---
title: "Practical Machine Learning - Project 01"
author: "Siva S."
date: "Sunday, March 22, 2015"
output: html_document
---

#Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

#Data that is being sampled

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. Thanks to the data provider for sharing the information for this project analsysis.

# Initial Analysis

Pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used.
During this project CARET and RANDOMFOREST packages were downloaded and installed.

# Model Details

Our outcome variable is classe, a factor variable with 5 levels. For this data set, “participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

* exactly according to the specification (Class A)
* throwing the elbows to the front (Class B)
* lifting the dumbbell only halfway (Class C)
* lowering the dumbbell only halfway (Class D)
* throwing the hips to the front (Class E)

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes." [1]
Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

# Model Validation

Validation is performed randomly on two data sets subTraining data (75% of the original Training data set) and subTesting data (25%).Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

# Expected sample error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

# Reasons for choosing this model

Our outcome variable “classe” is an unordered factor variable. Thus, we can choose our error type as 1-accuracy. We have a large sample size with N= 19622 in the Training data set. This allow us to divide our Training sample into subTraining and subTesting to allow cross-validation. Features with all missing values will be discarded as well as features that are irrelevant. All other features will be kept as relevant variables.
Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification [2]. Feature selection is inherent, so it is not so necessary at the data preparation phase. Thus, there won’t be any feature selection section in this report.

#Final CODE and SUMMARY RESULTS

### Load the required libraries
```{r}
  library(caret)
  library(randomForest)
  library(rpart)
  library(rpart.plot)
```

# Data Loading

```{r}
  library(RCurl)
  library(bitops)
  trainingset <- read.csv(file.choose(),na.strings=c("NA","#DIV/0!", "") )
  testingset <- read.csv(file.choose(), na.strings=c("NA","#DIV/0!", ""))
dim(trainingset)
  dim(testingset)
  trainingset<-trainingset[,colSums(is.na(trainingset)) == 0]
  testingset <-testingset[,colSums(is.na(testingset)) == 0]
  trainingset   <-trainingset[,-c(1:7)]
  testingset <-testingset[,-c(1:7)]
  dim(trainingset)
  dim(testingset)
  head(trainingset)
  head(testingset)
  
```

# Plot the data

```{r}
subsamples <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
subTraining <- trainingset[subsamples, ] 
subTesting <- trainingset[-subsamples, ]
dim(subTraining)
dim(subTesting)
head(subTraining)
head(subTesting)
plot(subTraining$classe, col="green", main="Bar Plot of levels subTraining data set", xlab="classe levels", ylab="Frequency")
```

# First Prediction with Decision Tree

```{r}
model1 <- rpart(classe ~ ., data=subTraining, method="class")

prediction1 <- predict(model1, subTesting, type = "class")

rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

```{r}
# You also need to have package e1071
confusionMatrix(prediction1, subTesting$classe)
```

#Second prediction model: Using Random Forest

```{r}
model2 <- randomForest(classe ~. , data=subTraining, method="class")

prediction2 <- predict(model2, subTesting, type = "class")

confusionMatrix(prediction2, subTesting$classe)

```

# Final Decision

As per the results the RANDOM FOREST algorithm fared better than DECISION TREE

# Final Submission

```{r}
predictfinal <- predict(model2, testingset, type="class")
predictfinal
```

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)

```

